Model-based Clustering for Neural Populations

### 2021-09-12 documents/cluster_neural_population.tex
modify TODO
transfer for debug dc MH

Since the progress noise is independent at each step,
\(f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right) = \prod_{t = 1}^{T}{P(y_{\text{it}}|\mathbf{\Theta}_{j})}\),
where \(P( \cdot )\) is the Poisson density and \(\mathbf{\Theta}_{j}\) contains all parameters in cluster \(j\).

@@

\section{TODO}
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Think about MV-(Poisson)-GLM. Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
	\item
	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
	\item
	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
	\item
	improve DPMM. In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)). There are several improvements, e.g. \href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}. Check them later.
	\item
	After all of them are resolved, switch to GP version

>>>>

In summary, Let
\(\mathbf{\Theta} = \left\{ \mathbf{z},\ \mathbf{d},\mathbf{C},\left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{x}_{0},\mathbf{Q}_{0},\mathbf{A},\ \mathbf{b},\mathbf{Q}\  \right\}\)
be the set of parameters for LDS. The number of observation/ neuron is
\(N\), and they can group into \(J\) clusters. In each cluster, there
are \(p\) latent state vectors. The recording length is \(T\).
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	\(\mathbf{z} \in \mathbb{Z}^{N}\): cluster index for each neuron/ observation,
	with \(z_{i} \in \{ 1,\ldots,J\}\) for \(i = 1,\ldots,N\).
	\item
	\(\mathbf{d} \in \mathbb{R}^{N}\) and \(\mathbf{C} \in \mathbb{R}^{N \times p}\): baseline
	\& loading of the latent
	\item
	\(\mathbf{x}_{t} \in \mathbb{R}^{\text{Jp}}\): all latent at \(t\).
	\item
	\(\mathbf{x}_{t}^{(j)} \in \mathbb{R}^{p}\) means the latent in cluster \(j\) at
	time \(t\).
	\item
	\(\mathbf{x}_{0} \in \mathbb{R}^{\text{Jp}}\) and
	\(\mathbf{Q}_{0} \in \mathbb{R}^{\text{Jp} \times Jp}\): mean and covariance of
	\(\mathbf{x}_{1}\). For simplicity, assume \(\mathbf{Q}_{0}\) is known.
	\item
	\(\mathbf{A} \in \mathbb{R}^{Jp \times Jp},\ \mathbf{b} \in \mathbb{R}^{Jp},\mathbf{Q} \in \mathbb{R}^{Jp \times Jp}\):
	linear dynamics of the latent.
\end{enumerate}

Since the progress noise is independent at each step and the observation is assumed conditional independent, the likelihood is:
\[ f(\mathbf{y}|\mathbf{\Theta})= \prod_{i = 1}^{N}\prod_{t = 1}^{T}{P(y_{it}|\mathbf{\Theta})} = \prod_{i = 1}^{N}\prod_{t = 1}^{T}{POI(y_{it}|\exp{(d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})})})}
\]
, where \(POI(\cdot|\lambda)\) is the density of \(Poisson(\lambda)\).

\textbf{DETOUR}:\\
Maybe in the future, we may switch to EM. To help with this, I also give the likelihood for complete observation \((\mathbf{y}, \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T})\). Let \(\mathbf{\Theta}'=\mathbf{\Theta}\backslash \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T}\).
\[ f(\mathbf{y}, \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T}|\mathbf{\Theta}')= \prod_{i = 1}^{N}\prod_{t = 1}^{T}{P(y_{it}, \mathbf{x}_{t}|\mathbf{\Theta}')} = \prod_{i = 1}^{N}\prod_{t = 1}^{T}{POI(y_{it}|\exp{(d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})})})}\cdot N(\mathbf{x}_{t}^{(z_{i})} |\mathbf{A}_{z_i}\mathbf{x}_{t} + \mathbf{b}_{z_i}, \mathbf{Q}_{z_i})
\]
, where \(\mathbf{A}_j\in \mathbb{R}^{p \times Jp}\) and \(\mathbf{b}_j\in \mathbb{R}^p\)  are the \(j^{\text{th}}\) row block of \(\mathbf{A}\) and \(\mathbf{b}\). \(\mathbf{Q}_j \in \mathbb{R}^{p \times p}\) is the \(j^{\text{th}}\) row and column block of \(\mathbf{Q}\).

@@

\section{TODO}
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Think about MV-(Poisson)-GLM. Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
	\item
	After resolving (1), see if the MCMC is still trapped in local optima sometimes. Now, the chain seems will get stuck in local mode. Maybe updating the loading as a whole will remedy the problem a bit.
	\item
	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
	\item
	improve DPMM. In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)). There are several improvements, e.g. \href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}. Check them later.
	\item
	After all of them are resolved, switch to GP version
